\proglang{Stan}: A Probabilistic Programming Language

### 2013-12-29 src/docs/papers/stan-language/article.tex
submitted version with pdf of paper and tarball of source

\renewcommand{\textfraction}{0.07}
\renewcommand{\topfraction}{0.9}	% max fraction of floats at top

@@

%% almost as usual
\author{{\bf\large Bob Carpenter}
        \\ Columbia University
        %
        \\[9pt]
        %
        {\bf\large Daniel Lee}
        \\ Columbia University
        %
        \\[9pt]
        %
        {\bf\large Marcus A.\ Brubaker}
        \\ TTI-Chicago
        %
    \And
        %
        {\bf\large Andrew Gelman}
        \\ Columbia University
        %
        \\[9pt]
        %
        {\bf\large Ben Goodrich}
        \\ Columbia University
        %
        \\[9pt]
        %
        {\bf\large Jiqiang Guo}
        \\ Columbia Univesity
        %
     \And
        %
        {\bf\large Matt Hoffman}
        \\ Adobe Research Labs
        %
        \\[9pt]
        %
        {\bf\large Michael Betancourt}
        \\ University College London
        %
        \\[9pt]
        %
        {\bf\large Peter Li}
        \\ Columbia University
}
\title{\proglang{Stan}: A Probabilistic Programming Language}

@@

  \proglang{Stan} is also a platform for computing log-densities and
  their gradients and Hessians, which can be used in alternative
  algorithms such as variational Bayes, expectation propagation, and
  marginal inference using approximate integration.  To this end, Stan
  is set up so that the densities, gradients, and Hessians, along with
  intermediate quantities of the algorithm such as acceptance
  probabilities, are easily accessible.

@@

These models turned out to be a challenge for
existing general-purpose inference software.  A direct encoding in
\proglang{BUGS} \citep{LunnEtAl:2000,LunnEtAl:2009,LunnEtAl:2012} or
\proglang{JAGS} \citep{Plummer:2003} can grind these tools to a halt.
We began by adding custom vectorized logistic regressions to \proglang{JAGS}
using \proglang{C++} to overcome the cost of interpretation.  Although
this is much faster than looping in \proglang{JAGS}, it quickly became
clear that the root of the problem was the slow convergence of
conditional sampling when parameters were highly correlated 
in the posterior, as for time-series models and hierarchical models 
with interacted predictors.  We finally realized we needed a better sampler, 
not a more efficient implementation of Gibbs sampling.

@@

We were hearing more about Hamiltonian Monte Carlo (HMC) sampling,
which appeared promising but was also problematic in that the
Hamiltonian dynamics simulation requires the gradient of the log
posterior.  Although not difficult in theory for most functions,
computing these gradients by hand on a model-by-model basis is very
tedious and error prone.  That is when we discovered reverse-mode
algorithmic differentiation, which, given a templated \proglang{C++}
function for the log posterior, automatically computes a proper
analytic gradient up to machine precision accuracy in only a few
multiples of the cost to evaluate the log probability function itself.
We explored existing algorithmic differentiation packages with open
licenses such as \pkg{RAD} \citep{Gay:2005} and its repackaging in the
Sacado module of the \pkg{Trilinos} toolkit, and the \pkg{CppAD}
package of the COIN-OR toolkit \citep{BellBurke:2008}.  Both
packages, however, supported few special functions (e.g., probability
functions, log gamma, inverse logit) or linear algebra operations
(e.g., Cholesky decompositions, matrix division), and neither are
easily or modularly extensible.

@@

At this point, we could fit models coded directly in \proglang{C++} on
top of the pre-release versions of the \proglang{Stan} API
(application programming interface).  Seeing how well this all worked,
we set our sights on the generality and ease of use of \proglang{BUGS}
and designed a modeling language in which statisticians could write
their models in familiar notation that could be transformed to
efficient \proglang{C++} code before being compiled into an efficient
executable program.  Although we started with the target of specifying
directed graphical models, our first implementation turned out to be a
more general imperative form of specifying log probability functions;
\proglang{BUGS} models could be translated line for line, but the
\proglang{Stan} language also supports much richer imperative
constructs like conditionals, while loops, and local variables.
This paper is primarily about the \proglang{Stan} language and how
it can be used to specify statistical models.

@@

Even with the prototype compiler generating models, we still faced a
major hurdle for ease of use. The efficiency of HMC is very sensitive
to two tuning parameters, the discretization interval (i.e., step
size) and the total simulation time (i.e., number of steps).  The
interval size parameter could be tuned during warm-up based on
Metropolis rejection rates, but the number of steps proved difficult
to tune without sacrificing the detailed balance of the sampler.  This
led to the development of the No U-Turn (NUTS) sampler
\citep{HoffmanGelman:2011}.  Roughly speaking, NUTS builds a tree of
possible samples by randomly simulating Hamiltonian dynamics both
forwards and backwards in time until the combined trajectory turns
back on itself.  Once the trajectory has terminated, a new sample is
drawn from the tree.

We thought we were home free at this point, but when we measured the
speed of some \proglang{BUGS} examples versus \proglang{Stan} we were
very disappointed.  \proglang{BUGS}'s very first example model, Rats,
ran more than an order of magnitude faster in \proglang{JAGS} than in
\proglang{Stan}.  Rats is a tough test case because the conjugate
priors and lack of posterior correlations make it an ideal candidate
for efficient Gibbs sampling.  Realizing that we were doing redundant
calculations, we wrote a vectorized form of the normal distribution
for multiple variates with the same mean and scale, which sped things
up a bit. At the same time, we introduced some simple template
metaprograms to remove the calculation of constant terms in the log
probability.  Performance was lacking until we finally figured out how
to both eliminate redundant calculations and partially evaluate the
gradients using a combination of expression templates and
metaprogramming.

When we attempted to fit a time-series model, we found that
normalizing the data to unit sample mean and variance sped up the fits
by an order of magnitude.  In hindsight this isn't surprising, as the
performance of the numerical simulation in HMC scales with the
variation of the parameter scales; fortunately HMC also provides a
choice of metric, in our case a Euclidean metric with a constant mass
matrix, to compensate for variations in the log posterior.  In
\proglang{Stan} 1.0 we introduced an adaptive diagonal metric into our
HMC implementations that allowed the parameter scales to be normalized
automatically, with the option of an adaptive dense metric and a full
automated standardization of the posterior in \proglang{Stan} 2.0.

Because the dense metric performs only a global standardization, the
performance of \proglang{Stan} in models where the relative parameter
scales varies by location, such as hierarchical and latent models, can
suffer.  Riemannian manifold Hamiltonian Monte Carlo (RHMC) introduces
a location-dependent metric that can overcome these final hurdles
\citep{GirolamiCalderhead:2011}; Stan has a prototype implementation
of RHMC based on the SoftAbs metric of \citep{Betancourt:2012}, using
a generalization of NUTS to Riemannian manifolds \citep{Betancourt:2013}.

@@

The executable can be run with default options by specifying a path to
the data file.  The first command in the following example changes the
current directory to that containing the model, which is where the
data resides and where the executable is built.  From there, the path
to the data is just the file name \code{bernoulli.data.R}.
%
\begin{CodeChunk}
\begin{CodeInput}
> cd $stan/src/models/basic_estimators
> ./bernoulli sample data file=bernoulli.data.R
\end{CodeInput}
\end{CodeChunk}
%
For Windows, the \code{./} before the command should be removed.  This
invocation specifies the executable program, the data file, the file
to which the samples are written, the identity of the chain, and a
seed for the random number generator.

@@

A description of all configuration parameters including default values
and constraints is available by executing
%
\begin{CodeChunk}
\begin{CodeInput}
> ./bernoulli help-all
\end{CodeInput}
\end{CodeChunk}
%
The sampler and its configuration are described at greater length in
the manual.

@@

The output CSV file, written by default to \code{output.csv}, starts
with a summary of the configuration parameters for the run.
%
\begin{Code}
# stan_version_major = 2
# stan_version_minor = 1
# stan_version_patch = 0
# model = bernoulli_model
# method = sample (Default)
#   sample
#     num_samples = 1000 (Default)
#     num_warmup = 1000 (Default)
#     save_warmup = 0 (Default)
#     thin = 1 (Default)
#     adapt
#       engaged = 1 (Default)
#       gamma = 0.050000000000000003 (Default)
#       delta = 0.80000000000000004 (Default)
#       kappa = 0.75 (Default)
#       t0 = 10 (Default)
#       init_buffer = 75 (Default)
#       term_buffer = 50 (Default)
#       window = 25 (Default)
#     algorithm = hmc (Default)
#       hmc
#         engine = nuts (Default)
#           nuts
#             max_depth = 10 (Default)
#         metric = diag_e (Default)
#         stepsize = 1 (Default)
#         stepsize_jitter = 0 (Default)
# id = 0 (Default)
# data
#   file = bernoulli.data.R
# init = 2 (Default)
# random
#   seed = 847896134
# output
#   file = output.csv (Default)
#   diagnostic_file =  (Default)
#   refresh = 100 (Default)
\end{Code}
%
\proglang{Stan}'s behavior is fully specified by these configuration
parameters.  By using the same version of \proglang{Stan} and these
configuration parameters, exactly the same output file can be
reproduced.  The pseudo-random numbers generated by the sampler are
fully determined by the seed (here randomly generated based on the
time of the run, with value \code{847896134}) and the identifier (here
\code{0}).  The identifier is used to advance the underlying
pseudo-random number generator a sufficient number of values that
using multiple chains with the same seed and different identifiers
will draw from different subsequences of the pseudo-random number
stream determined by the seed.

The output contiues with a CSV header naming the columns of the
output. For the default NUTS sampler in Stan 2.1.0, these are
%
\begin{Code}
lp__,accept_stat__,stepsize__,treedepth__,n_divergent__,theta
\end{Code}
%
The values for \code{lp\_\_} indicate the log probability (up to an
additive constant).  The column headed by \code{accept\_stat\_\_}
provides the Metropolis/slice acceptance statistic for each
iteration.%
%
\footnote{Acceptance is the usual notion for a Metropolis sampler such
  as HMC \citep{MetropolisEtAl:1953}.  For NUTS, the acceptance
  statistic is defined as the average acceptance probabilities of all
  possible samples in the proposed tree; NUTS itself uses a slice
  sampling algorithm for rejection \citep{Neal:2003}.}
%
The column \code{stepsize\_\_} indicates the step size
(i.e., time interval) of the simulated trajectory, while the column 
\code{treedepth\_\_} gives the tree depth for NUTS, defined as the 
log base 2 of the total number of steps in the trajectory.
The rest of the header will be the names of parameters; in this
example, \code{theta} is the only parameter.

Next, the results of adaptation are printed as comments.
%
\begin{Code}
# Adaptation terminated
# Step size = 0.783667
# Diagonal elements of inverse mass matrix:
# 0.517727
\end{Code}
%
By default, \proglang{Stan} uses the NUTS sampler with a diagonal mass
matrix.  The mass matrix is estimated, roughly speaking, by
regularizing the sample covariance of the latter half of the warmup
samplers; see \citep{Stan:2013} for full details.  A dense mass matrix
may also be estimated, or the mass matrix may be set to the unit matrix.

@@

\proglang{Stan} supplies a command-line program \code{bin/print} to
summarize the output of one or more MCMC chains.  Given a directory
containing output from sampling, it is invoked as
%
\begin{CodeChunk}
\begin{CodeInput}
> ls output*.csv
\end{CodeInput}
\begin{CodeOutput}
output.csv	output1.csv	output2.csv	output3.csv
\end{CodeOutput}
\begin{CodeInput}
> $stan/bin/print output*.csv
\end{CodeInput}

\end{CodeChunk}
%
The output is shown in Figure~\ref{print-output.fig}.%
%
\footnote{Aligning columns when printing rows of varying scales
  presents a challenge.  For each column, the program calculates the
  the maximum number of digits required to print an entry in that
  column with the specified precision. For example, a precision of 2
  for the number -0.000012 requires nine characters (\code{-0.000012})
  to print without scientific notation versus seven digits with
  (\code{-1.2e-5}).  If the discrepancy is above a fixed threshold,
  scientific notation is used.  Compare the results in the \code{mean}
  column versus the \code{sd} column.}
%

@@

                N_Eff  N_Eff/s     R_hat
lp__             1404    19447  1.00e+00
accept_stat__     887    12297  1.02e+00
stepsize__       2.00     27.7  5.56e+13
treedepth__      2856    39572  1.01e+00
n_divergent__    4000    55424       nan
theta            1399    19382  1.00e+00
\end{Code}
\caption{\it Output of \code{bin/print} for the Bernoulli estimation model in
  Figure~\ref{bernoulli-model.fig}.}\label{print-output.fig}
\end{figure}
%
Each row of the output summarizes a different value whose name is
provided in the first column.  These correspond to the columns in the
output CSV files. The analysis includes estimates of the posterior
mean (\code{Mean}) and standard deviation (\code{StdDev}).  The median
(\code{50\%}) and 90\% posterior interval (\code{5\%}, \code{95\%})
are also displayed.

@@

The column \code{N\_eff} is the number of effective samples in a
chain.  Because MCMC methods produce correlated samples in each chain,
estimates such as posterior means are not as accurate as they would be
with truly independent samples.  To number of effective samples is an
estimate of the number of independent samples that would lead to the
same accuracy.  The Monte Carlo standard error (\code{MCSE}) is an
estimate of the error in estimating the posterior mean based on
dividing the posterior standard deviation estimate by the square root
of the number of effective samples (\code{sd / sqrt(n\_eff)}).
\cite{Geyer:2011} provides a thorough introduction to effective sample
size and MCSE estimation.  \proglang{Stan} uses the more conservative
estimates based on both within-chain and cross-chain convergence; see
\citep{GelmanEtAl:2013} and \citep{Stan:2013} for motivation and
definitions.

@@

\subsection{Posterior Mode Estimates}

@@

 The optimizer
terminates when any of the log probability, gradient, or parameter
values are within their specified tolerance.  The default optimizer is
the Broyden-Fletcher-Goldfarb-Shanno (BFGS), a quasi-Newton method
which employs exactly computed gradients and an efficient
approximation to the Hessian; see \citep{NocedalWright:2006} for a
textbook exposition of the BFGS algorithm.

@@

Optimization is carried out on the unconstrained parameter space, but 
without the Jacobian adjustment to the log probability.

@@

\subsection{Example: hierarchical model, with inference}

@@

The \proglang{BUGS} surgical example explored the ranking of
institutions in terms of surgical mortality.  This is coded in the
example in Figure~\ref{hier-binom.fig} using the generated quantities
block.  The generated quantity variable \code{rnk[j]} will hold the
rank of institution \code{j} from 1 to \code{J} in terms of mortality
rate \code{theta[j]}.  The ranks are extracted using the \code{rank}
function. The posterior summary will print average rank and deviation,
but the values can also be extracted in order to plot posterior rank
histograms by institution as done in \citep[Examples, Volume
1]{LunnEtAl:2000}.

@@

A table summarizing the point at which variables are read, written,
or defined is provided in Figure~\ref{block-actions.fig}. 
%
\begin{figure}
\begin{center}
\begin{tabular}{l|c|l|l}
{ Block} & { Statements?} & { Action} & { Evaluated}
\\\hline\hline
\code{user initialization} & n/a & transform & chain
\\[3pt]
\code{random initialization} & n/a & randomize & chain 
\\\hline\hline
\code{data} & no & read & chain  
\\
\code{transformed data} & yes & evaluate & chain  
\\ \hline
\code{parameters} & no & inv.\ transform, Jacobian & leapfrog  \\
& & inv.\ transform, write & sample 
\\[3pt]
\code{transformed parameters} & yes & evaluate & leapfrog \\
& & write & sample 
\\\hline
\code{model} & yes & evaluate & leapfrog
\\\hline
\code{generated quantities} & yes & evaluate & sample \\
& & write & sample
\end{tabular}
\end{center}
\caption{\it Each \proglang{Stan} program block admits certain actions that are evaluated
  at specific times during sampling.  For example, the parameter initialization in
  the \code{data} block requires a read operation once per chain.}
\label{block-actions.fig}
\end{figure}
%
This table is defined assuming HMC or NUTS samplers, which require a
log probability and gradient calculation for one or more leapfrog
steps per iteration; see \citep{HoffmanGelman:2011}.
%
\begin{figure}
\begin{center}
\begin{tabular}{l|l}
{ Variable Categorization} & { Declaration Block}
\\ \hline\hline
% constants & \code{transformed data}
% \\ \hline
unmodeled data & \code{data}, \code{transformed data}
\\ 
modeled data & \code{data}, \code{transformed data}
\\ \hline
missing data & \code{parameters}, \code{transformed parameters}
\\
modeled parameters & \code{parameters}, \code{transformed parameters}
\\
unmodeled parameters & \code{data}, \code{transformed data}
\\[2pt] \hline
generated quantities & \code{transformed data}, \code{transformed parameters}, 
\\ 
& \code{generated quantities}
\\ \hline\hline
loop indices & any loop statement
\\ \hline
local variables & any statement block
\\ 
\end{tabular}
\end{center}
\caption{\it Variables of each categorization must be declared in
  specific blocks.  Data may also be expressed using numeric
  literals.}\label{variable-kinds.fig}
\end{figure}
%
\cite[p.~366]{GelmanHill:2007} provide a taxonomy of the kinds of
variables used in Bayesian models.  Figure~\ref{variable-kinds.fig}
contains Gelman and Hill's taxonomy aligned with the corresponding
locations of declarations and definitions in \proglang{Stan}.
Unmodeled data variables includes size constants and regression
predictors.  Modeled data variables include known outcomes or
measurements.  A data variable or constant literal is modeled in a
\proglang{Stan} program if it (or a variable that depends on it)
occurs on the left-hand side of a sampling statement.  Unmodeled
parameters are provided as data, and are either known or fixed to some
value for convenience, a typical use being the parameters of a weakly
informative prior for a parameter.

@@

\subsection{Implicit Uniform Priors}\label{implicit-prior.section}

>>>>

\renewcommand{\textfraction}{0.02}
\renewcommand{\topfraction}{0.95}	% max fraction of floats at top

@@

%% almost as usual
\author{{\bf\large Bob Carpenter}
        \\ Columbia University
        %
        \\[9pt]
        %
        {\bf\large Daniel Lee}
        \\ Columbia University
        %
        \\[9pt]
        %
        {\bf\large Marcus A.\ Brubaker}
        \\ TTI-Chicago
        %
        \\[9pt]
        %
        {\bf\large Allen Riddell}
        \\ Dartmouth College
    \And
        %
        {\bf\large Andrew Gelman}
        \\ Columbia University
        %
        \\[9pt]
        %
        {\bf\large Ben Goodrich}
        \\ Columbia University
        %
        \\[9pt]
        %
        {\bf\large Jiqiang Guo}
        \\ Columbia Univesity
        %
     \And
        %
        {\bf\large Matt Hoffman}
        \\ Adobe Research Labs
        %
        \\[9pt]
        %
        {\bf\large Michael Betancourt}
        \\ University College London
        %
        \\[9pt]
        %
        {\bf\large Peter Li}
        \\ Columbia University
        %
        \\[9pt]
}
\title{\proglang{Stan}: A Probabilistic Programming Language}

@@

  \proglang{Stan} is also a platform for computing log densities and
  their gradients and Hessians, which can be used in alternative
  algorithms such as variational Bayes, expectation propagation, and
  marginal inference using approximate integration.  To this end, Stan
  is set up so that the densities, gradients, and Hessians, along with
  intermediate quantities of the algorithm such as acceptance
  probabilities, are easily accessible.

@@

These models turned out to be a challenge for
existing general purpose inference software.  A direct encoding in
\proglang{BUGS} \citep{LunnEtAl:2000,LunnEtAl:2009,LunnEtAl:2012} or
\proglang{JAGS} \citep{Plummer:2003} can grind these tools to a halt.
We began by adding custom vectorized logistic regressions to \proglang{JAGS}
using \proglang{C++} to overcome the cost of interpretation.  Although
this is much faster than looping in \proglang{JAGS}, it quickly became
clear that the root of the problem was the slow convergence of
conditional sampling when parameters were highly correlated 
in the posterior, as for time-series models and hierarchical models 
with interacted predictors.  We finally realized we needed a better sampler, 
not a more efficient implementation of Gibbs sampling.

@@

We were hearing more about Hamiltonian Monte Carlo (HMC) sampling,
which appeared promising but was also problematic in that the
Hamiltonian dynamics simulation requires the gradient of the log
posterior.  Although not difficult in theory for most functions,
computing these gradients by hand on a model-by-model basis is very
tedious and error prone.  That is when we discovered reverse-mode
algorithmic differentiation, which, given a templated \proglang{C++}
function for the log posterior, automatically computes its analytic
gradient up to machine precision accuracy in only a few multiples of
the time to evaluate the log probability function itself.%
%
\footnote{Hessian matrices (i.e., all second order derivatives) are
  more expensive to calculate; each row corresponds to a parameter and
  is filled by the gradient of the derivative of the log probability
  function with respect to the parameter.}
%
We explored existing algorithmic differentiation packages with open
licenses such as \pkg{RAD} \citep{Gay:2005} and its repackaging in the
Sacado module of the \pkg{Trilinos} toolkit, and the \pkg{CppAD}
package of the COIN-OR toolkit \citep{BellBurke:2008}.  Both packages,
however, supported few special functions (e.g., probability functions,
log gamma, inverse logit) or linear algebra operations (e.g., Cholesky
decompositions, matrix division), and neither are easily or modularly
extensible.

@@

At this point, we could fit models coded directly in \proglang{C++} on
top of the pre-release versions of the \proglang{Stan} application
programming interface (API).  Seeing how well this all worked, we set
our sights on the generality and ease of use of \proglang{BUGS} and
designed a modeling language in which statisticians could write their
models in familiar notation that could be transformed to efficient
\proglang{C++} code before being compiled into an efficient executable
program.  Although we started with the target of specifying directed
graphical models, our first implementation turned out to be a more
general imperative form of specifying log probability functions;
\proglang{BUGS} models could be translated line for line, but the
\proglang{Stan} language also supports much richer imperative
constructs like conditionals, while loops, and local variables.  This
paper is primarily about the \proglang{Stan} language and how it can
be used to specify statistical models.

@@

Even with the prototype compiler generating models, we still faced a
major hurdle for ease of use. The efficiency of HMC is very sensitive
to two tuning parameters, the discretization interval (i.e., step
size) and the total simulation time (i.e., number of steps).  The
interval size parameter could be tuned during warm-up based on
Metropolis rejection rates, but the number of steps proved difficult
to tune without sacrificing the detailed balance of the sampler.  This
led to the development of the No U-Turn (NUTS) sampler
\citep{HoffmanGelman:2011}.  Roughly speaking, NUTS builds a tree of
possible samples by randomly simulating Hamiltonian dynamics both
forwards and backwards in time until the combined trajectory turns
back on itself.  Once the trajectory has terminated, a new sample is
drawn from the tree.  In its original version, NUTS also estimates a
step size during warmup based on a target acceptance probability.

We thought we were home free at this point, but when we measured the
speed of some \proglang{BUGS} examples versus \proglang{Stan} we were
very disappointed.  \proglang{BUGS}'s very first example model, Rats,
ran more than an order of magnitude faster in \proglang{JAGS} than in
\proglang{Stan}.  Rats is a tough test case because the conjugate
priors and lack of posterior correlations make it an ideal candidate
for efficient Gibbs sampling.  Realizing that we were doing redundant
calculations, we wrote a vectorized form of the normal distribution
for multiple variates with the same mean and scale, which sped things
up a bit.  Performance was lacking until we finally figured out how
to both eliminate redundant calculations and partially evaluate the
gradients using a combination of expression templates and
metaprogramming.

When we attempted to fit a time-series model, we found that
normalizing the data to unit sample mean and variance sped up the fits
by an order of magnitude.  In hindsight this isn't surprising, as the
performance of the numerical simulation in HMC scales with the
variation of the parameter scales.  In \proglang{Stan} 1.0 we
introduced an adaptive diagonal metric (mass matrix) into our HMC
implementations that allowed the parameter scales to be standardized
automatically; Stan 2.0 added an option to estimate a dense metric
(mass matrix) and a full covariance standardization of the posterior.

Because the diagonal and dense metrics perform only a global
standardization, the performance of \proglang{Stan} in models where
the relative parameter scales varies by location, such as hierarchical
and latent models, can suffer (hierarchical modeling is covered in
Sections~\ref{hierarchical-modeling.section} and reparameterizations to speed it
up in \ref{centering.section}).  Riemannian manifold Hamiltonian Monte
Carlo (RHMC) introduces a location-dependent metric that can overcome
these final hurdles \citep{GirolamiCalderhead:2011}; Stan has a
prototype implementation of RHMC based on the SoftAbs metric of
\citep{Betancourt:2012}, using a generalization of NUTS to Riemannian
manifolds \citep{Betancourt:2013}.

@@

The executable can be run with default options by specifying a path to
the data file.  The first command in the following example changes the
current directory to that containing the model, which is where the
data resides and where the executable is built.  From there, the path
to the data is just the file name \code{bernoulli.data.R}.
%
\begin{CodeChunk}
\begin{CodeInput}
> cd $stan/src/models/basic_estimators
> ./bernoulli sample data file=bernoulli.data.R
\end{CodeInput}
\end{CodeChunk}
%
For Windows, the \code{./} before the command should be removed.

@@

A description of all configuration parameters including default values
and constraints is available by executing
%
\begin{CodeChunk}
\begin{CodeInput}
> ./bernoulli help-all
\end{CodeInput}
\end{CodeChunk}
%
The sampler and its configuration are described at greater length in
the manual \citep{Stan:2013}.

@@

The output CSV file, written by default to \code{output.csv}, starts
with a summary of the configuration parameters for the run.
%
\begin{Code}
# stan_version_major = 2
# stan_version_minor = 1
# stan_version_patch = 0
# model = bernoulli_model
# method = sample (Default)
#   sample
#     num_samples = 1000 (Default)
#     num_warmup = 1000 (Default)
#     save_warmup = 0 (Default)
#     thin = 1 (Default)
#     adapt
#       engaged = 1 (Default)
#       gamma = 0.050000000000000003 (Default)
#       delta = 0.80000000000000004 (Default)
#       kappa = 0.75 (Default)
#       t0 = 10 (Default)
#       init_buffer = 75 (Default)
#       term_buffer = 50 (Default)
#       window = 25 (Default)
#     algorithm = hmc (Default)
#       hmc
#         engine = nuts (Default)
#           nuts
#             max_depth = 10 (Default)
#         metric = diag_e (Default)
#         stepsize = 1 (Default)
#         stepsize_jitter = 0 (Default)
# id = 0 (Default)
# data
#   file = bernoulli.data.R
# init = 2 (Default)
# random
#   seed = 847896134
# output
#   file = output.csv (Default)
#   diagnostic_file =  (Default)
#   refresh = 100 (Default)
\end{Code}
%
\proglang{Stan}'s behavior is fully specified by these configuration
parameters.  By using the same version of \proglang{Stan} and these
configuration parameters, exactly the same output file can be
reproduced.  The pseudorandom numbers generated by the sampler are
fully determined by the seed (here randomly generated based on the
time of the run, with value \code{847896134}) and the identifier (here
\code{0}).  The identifier is used to advance the underlying
pseudorandom number generator a sufficient number of values that
using multiple chains with the same seed and different identifiers
will draw from different subsequences of the pseudorandom number
stream determined by the seed.

The output contiues with a CSV header naming the columns of the
output. For the default NUTS sampler in Stan 2.1.0, these are
%
\begin{Code}
lp__,accept_stat__,stepsize__,treedepth__,n_divergent__,theta
\end{Code}
%
The values for \code{lp\_\_} indicate the log probability (up to an
additive constant).  The column headed by \code{accept\_stat\_\_}
provides the Metropolis/slice acceptance statistic for each
iteration.%
%
\footnote{Acceptance is the usual notion for a Metropolis sampler such
  as HMC \citep{MetropolisEtAl:1953}.  For NUTS, the acceptance
  statistic is defined as the average acceptance probabilities of all
  possible samples in the proposed tree; NUTS itself uses a slice
  sampling algorithm for rejection \citep{Neal:2003, HoffmanGelman:2011}.}
%
The column \code{stepsize\_\_} indicates the step size
(i.e., time interval) of the simulated trajectory, while the column 
\code{treedepth\_\_} gives the tree depth for NUTS, defined as the 
log base 2 of the total number of steps in the trajectory.
The rest of the header will be the names of parameters; in this
example, \code{theta} is the only parameter.

Next, the results of adaptation are printed as comments.
%
\begin{Code}
# Adaptation terminated
# Step size = 0.783667
# Diagonal elements of inverse mass matrix:
# 0.517727
\end{Code}
%
By default, \proglang{Stan} uses the NUTS sampler with a diagonal mass
matrix.  The mass matrix is estimated, roughly speaking, by
regularizing the sample covariance of the latter half of the warmup
samples; see \citep{Stan:2013} for full details.  A dense mass matrix
may also be estimated, or the mass matrix may be set to the unit matrix.

@@

@@

                N_Eff  N_Eff/s     R_hat
lp__             1404    19447  1.00e+00
accept_stat__     887    12297  1.02e+00
stepsize__       2.00     27.7  5.56e+13
treedepth__      2856    39572  1.01e+00
n_divergent__    4000    55424       nan
theta            1399    19382  1.00e+00
\end{Code}
\caption{\it Output of \code{bin/print} for the Bernoulli estimation model in
  Figure~\ref{bernoulli-model.fig}.}\label{print-output.fig}
\end{figure}

\proglang{Stan} supplies a command-line program \code{bin/print} to
summarize the output of one or more MCMC chains.  Given a directory
containing output from sampling,
%
\begin{CodeChunk}
\begin{CodeInput}
> ls output*.csv
\end{CodeInput}
\begin{CodeOutput}
output.csv	output1.csv	output2.csv	output3.csv
\end{CodeOutput}
\end{CodeChunk}
%
posterior summaries are printed using
%
\begin{CodeChunk}
\begin{CodeInput}
> $stan/bin/print output*.csv
\end{CodeInput}
\end{CodeChunk}
%
The output is shown in Figure~\ref{print-output.fig}.%
%
\footnote{Aligning columns when printing rows of varying scales
  presents a challenge.  For each column, the program calculates the
  the maximum number of digits required to print an entry in that
  column with the specified precision. For example, a precision of 2
  for the number -0.000012 requires nine characters (\code{-0.000012})
  to print without scientific notation versus seven digits with
  (\code{-1.2e-5}).  If the discrepancy is above a fixed threshold,
  scientific notation is used.  Compare the results in the \code{mean}
  column versus the \code{sd} column.}
%
Each row of the output summarizes a different value whose name is
provided in the first column.  These correspond to the columns in the
output CSV files. The analysis includes estimates of the posterior
mean (\code{Mean}) and standard deviation (\code{StdDev}).  The median
(\code{50\%}) and 90\% posterior interval (\code{5\%}, \code{95\%})
are also displayed.

@@

The column \code{N\_eff} is the number of effective samples in a
chain.  Because MCMC methods produce correlated samples in each chain,
estimates such as posterior means are not as accurate as they would be
with truly independent samples.  The number of effective samples is an
estimate of the number of independent samples that would lead to the
same accuracy.  The Monte Carlo standard error (\code{MCSE}) is an
estimate of the error in estimating the posterior mean based on
dividing the posterior standard deviation estimate by the square root
of the number of effective samples (\code{sd / sqrt(n\_eff)}).
\cite{Geyer:2011} provides a thorough introduction to effective sample
size and MCSE estimation.  \proglang{Stan} uses the more conservative
estimates based on both within-chain and cross-chain convergence; see
\citep{GelmanEtAl:2013} and \citep{Stan:2013} for motivation and
definitions.

@@

\subsection{Posterior mode estimates}

@@

The optimizer terminates when any of the log probability, gradient, or
parameter values are within their specified tolerance.  The default
optimizer uses the Broyden-Fletcher-Goldfarb-Shanno (BFGS) algorithm,
a quasi-Newton method which employs exactly computed gradients and an
efficient approximation to the Hessian; see \citep{NocedalWright:2006}
for a textbook exposition of the BFGS algorithm.

@@

Optimization is carried out on the unconstrained parameter space, but
without the Jacobian adjustment to the log probability.  This ensures
modes are defined with respect to the constrained parameter space as
declared in the parameters block and used in the model specification.
The need to suppress the Jacobian to match the scaling of the declared
parameters highlights the sensitivity of posterior modes to parameter
transforms.

@@

\subsection{Example: hierarchical model, with inference}\label{hierarchical-modeling.section}

@@

The \proglang{BUGS} surgical example explored the ranking of
institutions in terms of surgical mortality \citep[Examples, Volume
1]{LunnEtAl:2000}.  This is coded in the example in
Figure~\ref{hier-binom.fig} using the generated quantities block.  The
generated quantity variable \code{rnk[j]} will hold the rank of
institution \code{j} from 1 to \code{J} in terms of mortality rate
\code{theta[j]}.  The ranks are extracted using the \code{rank}
function. The posterior summary will print average rank and deviation.
\citep{LunnEtAl:2000} illustrated posterior inference by plotting
posterior rank histograms.

@@

A table summarizing the point at which variables are read, written,
or defined is provided in Figure~\ref{block-actions.fig}. 
%
\begin{figure}
\begin{center}
\begin{tabular}{l|c|l|l}
{ Block} & { Statements?} & { Action} & { Evaluated}
\\\hline\hline
\code{user initialization} & n/a & transform & chain
\\[3pt]
\code{random initialization} & n/a & randomize & chain 
\\\hline\hline
\code{data} & no & read & chain  
\\
\code{transformed data} & yes & evaluate & chain  
\\ \hline
\code{parameters} & no & inv.\ transform, Jacobian & leapfrog  \\
& & inv.\ transform, write & sample 
\\[3pt]
\code{transformed parameters} & yes & evaluate & leapfrog \\
& & write & sample 
\\\hline
\code{model} & yes & evaluate & leapfrog
\\\hline
\code{generated quantities} & yes & evaluate & sample \\
& & write & sample
\end{tabular}
\end{center}
\caption{\it Each \proglang{Stan} program block admits certain actions that are evaluated
  at specific times during sampling.  For example, the parameter initialization in
  the \code{data} block requires a read operation once per chain.}
\label{block-actions.fig}
\end{figure}
%
This table is defined assuming HMC or NUTS samplers, which require a
log probability and gradient calculation for one or more leapfrog
steps per iteration; see \citep{HoffmanGelman:2011}.
%
\begin{figure}
\begin{center}
\begin{tabular}{l|l}
{ Variable Categorization} & { Declaration Block}
\\ \hline\hline
% constants & \code{transformed data}
% \\ \hline
unmodeled data & \code{data}, \code{transformed data}
\\ 
modeled data & \code{data}, \code{transformed data}
\\ \hline
missing data & \code{parameters}, \code{transformed parameters}
\\
modeled parameters & \code{parameters}, \code{transformed parameters}
\\
unmodeled parameters & \code{data}, \code{transformed data}
\\[2pt] \hline
generated quantities & \code{transformed data}, \code{transformed parameters}, 
\\ 
& \code{generated quantities}
\\ \hline\hline
loop indices & any loop statement
\\ \hline
local variables & any statement block
\\ 
\end{tabular}
\end{center}
\caption{\it Variables of each categorization must be declared in
  specific blocks.  Data may also be expressed using numeric
  literals.}\label{variable-kinds.fig}
\end{figure}
%
\citet[p.~366]{GelmanHill:2007} provide a taxonomy of the kinds of
variables used in Bayesian models.  Figure~\ref{variable-kinds.fig}
contains Gelman and Hill's taxonomy aligned with the corresponding
locations of declarations and definitions in \proglang{Stan}.
Unmodeled data variables includes size constants and regression
predictors.  Modeled data variables include known outcomes or
measurements.  A data variable or constant literal is modeled in a
\proglang{Stan} program if it (or a variable that depends on it)
occurs on the left-hand side of a sampling statement.  Unmodeled
parameters are provided as data, and are either known or fixed to some
value for convenience, a typical use being the parameters of a weakly
informative prior for a parameter.

@@

\subsection{Implicit uniform priors}\label{implicit-prior.section}
